---
title: "Audit, local runtime, privacy and troubleshooting"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Audit, local runtime, privacy and troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
example_root <- tempfile("cttir-audit-")
dir.create(example_root)
old_options <- options(cttiR.catalog_dir = file.path(example_root, "catalog"),
  cttiR.runtime_dir = file.path(example_root, "runtime"))
```

## doctor(), audit() and update() do different things

| Function | Purpose | Side effects |
|---|---|---|
| `doctor()` | brief read-only diagnostics | none |
| `audit()` | full check registry, reports, optional allowlisted repairs | reports only when `output` is given; repairs only with `repair = TRUE` |
| `update()` | refresh the knowledge/resource catalogs | new immutable snapshots and one pointer change; never installs packages |

```{r audit}
p <- cttiR::project("Audit example", "methods", "Describe a tabular comparison", path = example_root)
report <- cttiR::audit(p$path)
report
head(as.data.frame(report)[, c("id", "status", "message")], 10)
```

The default audit reads metadata only. It never opens referenced datasets,
runs study pipelines, installs dependencies, starts services or contacts a
network. The `integration` scope opts into temporary synthetic projects and an
isolated catalog fixture; `live = TRUE` additionally allows a bounded probe of an
already running owned local model.

Repairs are an explicit allowlist: restore a missing managed file whose
baseline is known, roll back an interrupted write whose writer has stopped, and
repoint a corrupt catalog pointer to the last verified snapshot. Edited files
are never overwritten.

```{r repair}
unlink(file.path(p$path, "code", "validate_project.R"))
cttiR::audit(p$path, scope = "project")$overall_status
repaired <- cttiR::audit(p$path, scope = "project", repair = TRUE)
repaired$overall_status
file.exists(file.path(p$path, "code", "validate_project.R"))
```

`audit(strict = TRUE)` raises `cttir_audit_failed` after writing any requested
reports when a required check fails or stays unverified; the condition carries
the full report for CI use.

## Readiness levels

Project creation always reports `scaffold_ready`. Higher levels come from local
evidence: matching installed or renv-locked dependency versions
(`environment_ready`), a passing synthetic demonstration produced by the
unmodified stage code (`demo_verified`), a passing study-data run receipt
(`data_ready`) and, finally, recorded mappings, reviewed settings, analysis
approval and approved stage adapters (`analysis_ready`). None of these is a
scientific validation of results.

## Local runtime and planner

`setup()` prepares an owned, cloud-disabled Ollama runtime on Linux x86_64 from
a pinned publisher archive and verifies a local model with a bounded structured
probe. A dry run is read-only:

```{r setup-plan}
plan <- cttiR::setup(dry_run = TRUE)
plan$state
plan$model$name
```

The default project planner is deterministic. Tested local models are listed
with their benchmark results in the runtime manifest; none met the
pre-registered planning thresholds. With `options(cttiR.planner = "local_llm")`
and a verified owned runtime, a model plans only if the manifest marks it
qualified at its exact digest; otherwise the deterministic planner is used and
the fallback reason `model_not_qualified` is recorded. Using an unqualified
model needs the separate acknowledgement
`options(cttiR.planner_allow_unqualified = TRUE)`, and its qualification label
is then stored with every decision it made. Goals that read like instructions
to the model are never sent (`goal_injection_suspected`). Explicit options
always win over suggestions, and replay or `sync()` never re-plan.

## Data flow and privacy

* Goals, configuration files, documentation and model output are treated as
  untrusted data; nothing is evaluated.
* Study data are never read during creation, synchronization, audit or catalog
  updates. `check_analysis_data()` and `code/run_workflow.R` read data only when
  you supply or bind them explicitly.
* Machine-specific paths belong in the ignored `.cttir/local.yml`.
* The planner (when enabled) and `ask()` never send text off the machine; the
  runtime listens on loopback only and refuses cloud-backed models.
* Exported Shiny drafts can contain your text; review them before sharing.

## Troubleshooting

| Symptom | Next step |
|---|---|
| `cttir_transaction_conflict` with an active writer | wait for the other operation; if it crashed, run `cttiR::audit(path, repair = TRUE)` |
| `cttir_catalog_corrupt` | `cttiR::audit(scope = "knowledge", repair = TRUE)` or `cttiR::rollback_knowledge(<id>, dry_run = FALSE)` |
| `workflow_approval_pending` blocker | the pinned catalog lacks approvals for a stage; review with `cttiR::packages(path)` |
| `environment_pending` blocker | install the pinned versions or enable `workflow$environment = "renv"` with `prepare_environment = TRUE` |
| `code/run_workflow.R` stops with a list | record the listed mappings/settings/approval with `cttiR::sync()` and bind the dataset in `.cttir/local.yml` |
| `setup()` returns `blocked` | read `$blockers`; offline mode never downloads, and only Linux x86_64 is supported |

```{r cleanup, include=FALSE}
options(old_options)
unlink(example_root, recursive = TRUE)
```
