---
title: "Audit, local runtime, privacy and troubleshooting"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Audit, local runtime, privacy and troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
example_root <- tempfile("cttir-audit-")
dir.create(example_root)
old_options <- options(cttiR.catalog_dir = file.path(example_root, "catalog"),
  cttiR.runtime_dir = file.path(example_root, "runtime"))
```

## doctor(), audit() and update() do different things

| Function | Purpose | Side effects |
|---|---|---|
| `doctor()` | brief read-only diagnostics | none |
| `audit()` | full check registry, reports, optional allowlisted repairs | reports only when `output` is given; repairs only with `repair = TRUE` |
| `update()` | refresh the knowledge/resource catalogs | new immutable snapshots and one pointer change; never installs packages |

```{r audit}
p <- cttiR::project("Audit example", "methods", "Describe a tabular comparison", path = example_root)
report <- cttiR::audit(p$path)
report
head(as.data.frame(report)[, c("id", "status", "message")], 10)
```

The default audit reads metadata only. It never opens referenced datasets,
runs study pipelines, installs dependencies, starts services or contacts a
network. The `integration` scope opts into temporary synthetic projects and an
isolated catalog fixture; `live = TRUE` additionally allows a bounded probe of an
already running owned local model.

`RES-008` compares every required and optional dependency pin with installed
version metadata, including CRAN interoperability packages. In renv mode it reads
the project library; a matching global installation does not substitute for a
missing project package. Unresolved pins warn for a scaffold and fail an
`analysis_ready` claim. Matching version numbers alone do not establish matching
source hashes or scientific suitability. Nothing is installed by this check.

Generated interoperability helpers also enforce pins at runtime. Run them from
the project root so they can read `cttir-lock.json`, or pass a complete reviewed
`pins` list explicitly. They refuse a missing lock, missing package pins and
installed-version mismatches, including when the class namespace is already
loaded. The lock is read as JSON; no project code is sourced, and nothing is
installed automatically. `jsonlite` reads the lock and its version is then
checked against that lock before object operations proceed.

Repairs are an explicit allowlist: restore a missing managed file whose
baseline is known, roll back an interrupted write whose writer has stopped, and
repoint a corrupt catalog pointer to the last verified snapshot. Edited files
are never overwritten.

```{r repair}
unlink(file.path(p$path, "code", "validate_project.R"))
cttiR::audit(p$path, scope = "project")$overall_status
repaired <- cttiR::audit(p$path, scope = "project", repair = TRUE)
repaired$overall_status
file.exists(file.path(p$path, "code", "validate_project.R"))
```

`audit(strict = TRUE)` raises `cttir_audit_failed` after writing any requested
reports when a required check fails or stays unverified; the condition carries
the full report for CI use.

## Readiness levels

Project creation always reports `scaffold_ready`. Higher levels come from local
evidence: matching installed or renv-locked dependency versions
(`environment_ready`), a passing synthetic demonstration produced by the
unmodified stage code (`demo_verified`), a passing study-data run receipt
(`data_ready`) and, finally, recorded mappings, reviewed settings, analysis
approval and approved stage adapters (`analysis_ready`). None of these is a
scientific validation of results.

## Local runtime and planner

`setup()` prepares an owned, cloud-disabled Ollama runtime on Linux x86_64 from
a pinned publisher archive and verifies a local model with a bounded structured
probe. A dry run is read-only:

```{r setup-plan}
plan <- cttiR::setup(dry_run = TRUE)
plan$state
plan$model$name
```

The default project planner is deterministic. The recorded 7B CPU results are
currently development evidence only. Fresh independent qualification on the
shipped artifact remains pending, so no manifest model authorizes automatic
planning. Automatic model selection retains its tested resource policy, but
blocks acquisition when no independently qualified model is available.
With `options(cttiR.planner = "local_llm")`
and a verified owned runtime, a model plans only if the manifest marks it
qualified at its exact digest and the tested prompt, grounding policy, registry,
schema, runtime binary and inference settings still match; otherwise the deterministic planner is used and
the fallback reason `model_not_qualified` is recorded. Using an unqualified
model needs the separate acknowledgement
`options(cttiR.planner_allow_unqualified = TRUE)`, and its qualification label
is then stored with every decision it made. Goals that read like instructions
to the model are never sent (`goal_injection_suspected`). Explicit options
always win over suggestions, and replay or `sync()` never re-plan.

The prompt includes registered package names and aliases. A validated model
proposal retains only capabilities with registered keyword evidence in the goal;
omissions are recorded in bounded notes and the project decision evidence.
This can omit legitimate requests expressed without those keywords. Filtering
is not workflow approval, and fixed inference settings do not guarantee identical
suggestions. Review proposed study decisions before using them for analysis.
Long inputs are bounded again after JSON escaping, preserving valid input JSON.
Planner 4 keeps the longest prefix that fits this character budget. Every reply,
including a repair, must report a valid prompt token count with room for the
reserved output tokens in the configured context. Missing or excessive counts
cause `context_overflow` fallback without another request. This checks the
runtime's reported count; it is not an independent tokenizer or proof that the
runtime preserved every input token. Earlier planner evaluations remain
historical development evidence and do not qualify the new policy.

## Data flow and privacy

* Goals, configuration files, documentation and model output are treated as
  untrusted data; nothing is evaluated.
* Study data are never read during creation, synchronization, audit or catalog
  updates. `check_analysis_data()` and `code/run_workflow.R` read data only when
  you supply or bind them explicitly.
* Machine-specific paths belong in the ignored `.cttir/local.yml`.
* The planner (when enabled) and `ask()` never send text off the machine; the
  runtime listens on loopback only and refuses cloud-backed models.
* Exported Shiny drafts can contain your text; review them before sharing.

## Troubleshooting

| Symptom | Next step |
|---|---|
| `cttir_transaction_conflict` with an active writer | wait for the other operation; if it crashed, run `cttiR::audit(path, repair = TRUE)` |
| `cttir_catalog_corrupt` | `cttiR::audit(scope = "knowledge", repair = TRUE)` or `cttiR::rollback_knowledge(<id>, dry_run = FALSE)` |
| `workflow_approval_pending` blocker | the pinned catalog lacks approvals for a stage; review with `cttiR::packages(path)` |
| `environment_pending` blocker | install the pinned versions or enable `workflow$environment = "renv"` with `prepare_environment = TRUE` |
| `code/run_workflow.R` stops with a list | record the listed mappings/settings/approval with `cttiR::sync()` and bind the dataset in `.cttir/local.yml` |
| `setup()` returns `blocked` | read `$blockers`; offline mode never downloads, and only Linux x86_64 is supported |

```{r cleanup, include=FALSE}
options(old_options)
unlink(example_root, recursive = TRUE)
```
