---
title: "Create, inspect and maintain a research scaffold"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Create, inspect and maintain a research scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# This vignette owns only this disposable example directory.
example_root <- tempfile("cttir-vignette-")
dir.create(example_root)
old_catalog_option <- getOption("cttiR.catalog_dir")
options(cttiR.catalog_dir = file.path(example_root, "catalog"))
```

cttiR creates a scaffold from a name, research type and goal. The current
implementation records unresolved scientific decisions and provides a reviewed
layout adaptation, catalog evidence and transactional maintenance. It does not
yet implement the complete research workflow or confer scientific approval.

## Start with three inputs

`path` is the existing **parent** directory. Creation uses a portable child name.
A preview does not create the child or install dependencies.

```{r preview}
plan <- cttiR::project(
  "Example comparison", "methods", "Plan a reproducible tabular comparison",
  path = example_root, dry_run = TRUE
)
plan$spec$project[c("name", "slug", "type")]
plan$readiness
stopifnot(!dir.exists(plan$path))
```

Apply the same request explicitly:

```{r create}
p <- cttiR::project(
  "Example comparison", "methods", "Plan a reproducible tabular comparison",
  path = example_root
)
stopifnot(dir.exists(p$path))
p$spec$analysis
```

Creation does not read study data, infer endpoints or ethics approval, call a
model, initialize Git, install packages or render the report. The configuration
and lock record accepted identity and provenance. The lock is not an `renv.lock`.

The project contains:

- `cttir-project.yml`: project, research and publication configuration.
- `cttir-lock.json`: accepted catalog, resource and template pins.
- `metadata/`: dataset registry, variable dictionary and template provenance.
- `protocol/analysis-plan.md`: decisions to review before analysis.
- `analysis/`: adapted reflowR pages and a labelled synthetic rendering fixture.
- `publications/`: separate analysis, manuscript, figure, table and submission trees.
- `.cttir/`: ownership, state and machine-local data bindings.

Raw data are not copied. Keep sensitive local paths in the ignored
`.cttir/local.yml`; registry descriptions should be suitable for the intended
repository audience.

## Repeat safely and inspect explicit changes

An identical creation request preserves the accepted identity and existing edits.
A different request for an occupied project name raises a conflict; use `sync()`
to preview configuration changes.

```{r repeat}
again <- cttiR::project(
  "Example comparison", "methods", "Plan a reproducible tabular comparison",
  path = example_root
)
stopifnot(identical(again$spec$project$id, p$spec$project$id))
stopifnot(all(again$plan$action == "skip"))
```

`options` takes precedence over a `config` list or YAML/JSON file. Unknown fields
are rejected. Updates to publication and dataset arrays are keyed by identity.
Unsupported workflow or dependency migrations fail explicitly.

```{r sync}
preview <- cttiR::sync(p$path, options = list(project = list(language = "de")))
preview$actions[preview$actions$action != "skip", c("path", "action")]
stopifnot(preview$state == "planned")
applied <- cttiR::sync(
  p$path, options = list(project = list(language = "de")), dry_run = FALSE
)
applied$state
```

The language setting is metadata; it does not promise that every generated page
or interface message is translated. User-owned analysis/manuscript files are
preserved. Edited managed files cause a conflict. Applied changes have backups and
a journal, but a multi-file update is not one atomic filesystem operation.

## Record analysis mappings without running a model

Dataset identifiers link the analysis to its registry entry. Column names are
literal names, including names containing spaces; they are never parsed as R
expressions. Unknown decisions stay visible. This example records an intended
continuous-outcome association and leaves approval unresolved:

```{r analysis-mapping}
analysis_config <- list(
  data_sources = list(list(
    id = "source01", label = "Example registry entry", logical_uri = NULL,
    format = "csv", access_class = "unknown", checksum = NULL,
    checksum_status = "unknown", schema_ref = NULL
  )),
  analysis = list(
    aim = "explanatory", outcome_family = "continuous",
    unit_structure = "independent", engine = "stats::lm", approved = FALSE,
    mapping = list(
      data_source_id = "source01", outcome = "response",
      predictors = list("exposure"), estimand = "A specified conditional association",
      missing_data = "fail"
    )
  )
)
mapped <- cttiR::project(
  "Mapping example", "methods", "Record a proposed analysis configuration",
  path = example_root, config = analysis_config, dry_run = TRUE
)
mapped$readiness$analysis[c("state", "candidate_engine", "missing_fields", "executable")]
stopifnot(!mapped$readiness$analysis$executable, !dir.exists(mapped$path))
```

Binary and survival configurations require explicit event/non-event codes;
survival also requires time origin and units. Longitudinal configurations require
subject/time mappings and a separate review of random effects and residual
structure. Predictive, causal and other unsupported designs remain capability
gaps. An explicitly empty predictor list records an intercept-only intention.
`complete_case` records a reviewed missing-data intention; planning never removes
rows. Even a fully recorded configuration and `approved = TRUE` cannot bypass
pending adapter/revision approval, dataset checks or model diagnostics.

Check an explicitly supplied in-memory data frame separately. This does not read
registered data locations, modify records or fit a model. Only mapped columns
participate in missingness; a configured complete-case policy reports how many
rows are complete without deleting them. Event codes are compared as literal
text, including factor labels. Data frame subclasses, matrix/list columns and
unreviewed date transformations are refused rather than coerced.

```{r explicit-data-check}
synthetic <- data.frame(response = c(2, 4, 6, 8), exposure = 1:4)
data_check <- cttiR::check_analysis_data(synthetic, mapped$spec)
data_check[c("state", "rows_total", "rows_complete", "issues", "executable")]
stopifnot(data_check$state == "passed", !data_check$executable)
```

A passed result covers only the listed structural checks. It does not establish
independence, unit correctness, rank, model assumptions, adequate sample size,
causal identification or scientific approval. Survival checks currently target
positive follow-up times with explicit right-censoring codes; more complex event
processes require a different adapter.

## Query evidence without approving a workflow

The API catalog and resource registry are separate evidence layers. A discovered
resource is not automatically installed or approved. An extracted function
signature is not proof of scientific suitability.

```{r evidence}
hit <- cttiR::search("reflowR::reflow_init", path = p$path)
hit[, c("package", "revision", "verification", "approved")]
answer <- cttiR::ask("reflowR::reflow_init", path = p$path)
stopifnot(identical(answer$code, ""))
```

Literal documentation search can return bounded excerpts from stored, licensed
source documents. The initial corpus includes the pinned reflowR revision.
Documents are not rendered or executed during indexing. Coverage is reported
separately from workflow approval:

```{r coverage}
known <- cttiR::packages(path = p$path)
known[known$package == "reflowR",
  c("package", "documents_discovered", "documents_stored", "approved")]
```

`ask(verified_only = TRUE)` restricts evidence to statically resolved API records.
It does not yet provide qualified natural-language planning. The package currently
has zero workflow-approved revisions, so a complete research-ready workflow must
not be inferred from these queries.

## Refresh only sources you select

The following examples are deliberately not run by the vignette. They show
explicit local and public GitHub source registration:

```{r update-examples, eval=FALSE}
options(cttiR.sources = list(
  list(id = "local-tool", path = "/path/to/package-source"),
  list(id = "public-reflow", github = "CTTIR/reflowR", package = "reflowR",
       ref = "cd1243a068ff2c8fb6796e34b58f6c6ce6af87e8")
))
cttiR::update(dry_run = TRUE)
cttiR::update(sources = "public-reflow", mode = "remote", dry_run = TRUE)
```

To store document text, a registration must also declare a reviewed
`documentation_rights` basis covering those documents. Without it, the updater
retains inventory and hashes only. Do not assume a package license covers all
third-party assets.

Remote GitHub mode resolves one commit and verifies bounded downloads against
Git blob hashes. It does not install the source package. Local mode never fetches
remote records. Discovery, pruning, non-GitHub backends and release migration remain
pending. A failed selected source prevents activation.

Updates retain complete immutable snapshots. Existing projects keep their pins;
removed exports do not reappear through fallback to an older bundled package.
Use `rollback_knowledge(id)` to preview a retained composite snapshot and
`dry_run = FALSE` to apply it. Initial resource observation times can make preview
and applied composite IDs differ even when API content is identical.

An applied update can recover a verified stopped local catalog writer after
checking the active snapshot and activation journal. It retains old writer
evidence. Live or unknown writers, conflicting journals and corrupt snapshots
are refused. An interrupted recovery guard requires manual review.

## Audit and render intentionally

Audit is local and read-only by default. It does not open referenced datasets or
run an analysis. This example checks only the project:

```{r audit}
report <- cttiR::audit(p$path, scope = "project")
report$overall_status
```

A warning or untested readiness check is not scientific clearance. `strict = TRUE`
raises a typed condition for required failures or required unverified checks.
`repair = TRUE` uses a restricted repair set: it does not overwrite arbitrary
user edits. Consult `help("audit", package = "cttiR")` before repair.

For new adapted-layout projects, run this from the generated project root only
after installing the rendering dependencies explicitly and reviewing any page edits:

```{r render-example, eval=FALSE}
# In a terminal at the generated project root:
# Rscript code/render_report.R
```

Rendering intentionally executes reviewed R Markdown chunks. The bundled fixture
uses synthetic values; it does not fit the project's study data. The adaptation
records its pinned reflowR source and MIT notice, while using
`rmarkdown::render_site` explicitly. It does not invoke `reflow_init()` or
workflowr and does not claim all upstream reproducibility features.

## Optional local interface and runtime

With `shiny` and `callr` installed, `setup_app()` provides Fast/Detailed creation,
and `configure(path)` uses the same synchronization engine. It is a local
interface. Use **Save or resume a draft** to export a JSON file or restore saved
answers. Drafts contain user text/configuration but exclude the parent directory,
local bindings and accepted preview. Restoring requires a fresh preview, and
Configure drafts must match the project identity. Review sensitive content before
sharing; drafts are not saved automatically. The full typed questionnaire and
remaining workflow views are still incomplete.

```{r optional-tools, eval=FALSE}
cttiR::setup_app()
cttiR::configure(p$path)
cttiR::setup(dry_run = TRUE)
```

Runtime setup is explicit. The live-tested installer currently targets Linux
x86_64 and verifies the pinned Ollama runtime and local model. Offline setup
never acquires software or pulls a model. A successful inference probe is not a
qualified workflow planner. No CRAN submission is performed by these functions.

```{r cleanup, include=FALSE}
options(cttiR.catalog_dir = old_catalog_option)
unlink(example_root, recursive = TRUE)
```
