---
title: "Knowledge catalog, approvals and grounded answers"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Knowledge catalog, approvals and grounded answers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
example_root <- tempfile("cttir-knowledge-")
dir.create(example_root)
old_options <- options(cttiR.catalog_dir = file.path(example_root, "catalog"), cttiR.sources = NULL)
```

cttiR ships an immutable, revision-scoped knowledge catalog. It records public
exports, statically resolved signatures, S3/S4/S7 declarations, documentation
inventories and reviewed **workflow approvals** for exact package revisions.
Generated project code may call only APIs covered by an approval of the
revision the project pins.

## What is cataloged

```{r packages}
inventory <- cttiR::packages()
nrow(inventory)
table(inventory$provider)
head(inventory[inventory$approved > 0, c("package", "version", "approved", "approved_roles")], 8)
```

CTTIR packages are indexed from verified public revisions. The standard
workflow family (tidyverse components, DescrTab2, ggplot2/patchwork, palettes,
model engines, broom, rendering, pipeline and environment tools, Bioconductor
containers and Seurat) is indexed from verified CRAN/Bioconductor source
archives and the R 4.6.1 distribution.

## Evidence levels

| Level | Meaning |
|---|---|
| `static_api_verified` | A literal function definition was parsed; nothing was executed. |
| `installed_api_verified` | Formals were read from the identical installed version in an isolated process (used where static parsing cannot resolve a constructor-built function). |
| `reexport_declared` | The export belongs to another package; it is approved only through its owner. |
| `static_declaration_only` | A literal S4/S7 declaration; dispatch is shown by fixtures, not by parsing. |
| `workflow_approved` | Covered by a complete reviewed approval of this exact revision. |

An approval names the package revision (`source_hash`), the workflow role and
adapter version, the required callables, S3 methods and documented objects, the
stored reference topics and package documents, and the passing fixture tests.
Any change to the source, including documentation only, produces a new
revision that needs a new review; older projects keep their pinned approvals.

```{r approvals}
approved <- cttiR::search("nlme::lme")
approved[, c("symbol", "verification", "approved")]
```

The bundled corpus stores the topics and package documents that approvals
require. Other documents of the same revision are inventoried with their hashes
as `indexed_not_bundled`; a local `cttiR::update()` with a reviewed
`documentation_rights` basis can store them.

## Grounded answers

`ask()` reads a question as data, matches it to registered capabilities and
exact `package::export` names, and returns steps, prerequisites, the approved
package revisions, citations and a reviewed snippet that validates against the
approved function table. It never runs code or calls a model.

```{r ask}
answer <- cttiR::ask("Fit a mixed model for repeated measures and get confidence intervals")
answer
answer$evidence[, c("package", "export", "verification", "approval")]
```

Unsupported procedures return a gap and the nearest approved alternatives
instead of invented code:

```{r ask-gap}
cttiR::ask("Run a Bayesian network meta-analysis with brms")$gaps
cttiR::ask("Does dplyr::old_removed_function exist?")$gaps
```

The bundled benchmark (`inst/benchmarks/ask-cases.json`) checks citation
correctness, code validity, abstention and injection resistance. Keyword
matching is safe but limited on unseen phrasing; the held-out results are
recorded in `inst/benchmarks/ask-heldout.json`.

## Updating and rolling back

`update()` refreshes registered sources into new immutable snapshots and
activates them with one pointer change; it never installs packages or changes
project pins. Registrations can be local directories, public GitHub
repositories, CRAN or Bioconductor packages (remote mode only) or a local R
distribution tree. Previews use disposable storage.

```{r update, eval = FALSE}
options(cttiR.sources = list(
  list(id = "cran-patchwork", cran = "patchwork", documentation_rights = "<reviewed basis>")
))
preview <- cttiR::update(mode = "remote", dry_run = TRUE)
preview$api_diff
preview$approval_diff
applied <- cttiR::update(mode = "remote")
cttiR::rollback_knowledge(applied$previous_id, dry_run = TRUE)
```

A refreshed revision whose source differs from an approved one keeps no
approval until it is reviewed; `approval_diff` reports such invalidations.
Removed exports never fall back to older bundled revisions.

```{r cleanup, include=FALSE}
options(old_options)
unlink(example_root, recursive = TRUE)
```
